Carbonate clumped isotope constraints on Silurian ocean temperature and \deltao O

### 2013-07-29 draft.tex
in draft.tex, included newest versions of figs, re-wrote fig captions, also made two column

\documentclass[preprint, authoryear]{elsarticle}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{rotating}
\usepackage{multirow}
\usepackage{enumerate}
\usepackage{wasysym}
\usepackage{natbib}
\usepackage{lineno}

@@

\linenumbers

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

Our results, along with those of previous clumped isotope studies \citep{Came2007,Dennis2013,Finnegan2011}, provide consistent explanations for the low \deltao O values of early Paleozoic carbonates (Figure \ref{Veizer}). Carbonate clumped isotopes constrain the contribution of a change in seawater \deltao O to only a few permil of the observed trend, ruling out a value of $-8\permil$ for the \deltao O$_{\text{SMOW}}$ of early Paleozoic oceans \citep{Grossman2012}. Our results also serve as a reminder that many of the fossils preserved in the geologic record lived in tropical oceans. The \deltao O of carbonate in equilibrium with tropical ocean water should not be compared to the \deltao O of carbonate in equilibrium with water of average modern sea surface temperature, because tropical oceans are much warmer than the average sea surface temperature. Clumped isotopes additionally demonstrate that extremely hot ($>50$\degrees C) ocean temperatures cannot be the cause of the depleted \deltao O values of early Paleozoic carbonates. Finally, clumped isotopes, as a highly sensitive indicator of diagenetic alteration, demonstrate that even among samples with original calcite fabric and pristine trace metal signatures, there is a large range in preservation quality. This means that the average fossil calcite in almost any sample suite is likely altered from its primary composition, and that diagenetic alteration is a key contributor to the low \deltao O values of early Paleozoic carbonates.

@@

The results from this study join those of previous clumped isotope studies in implying that the \deltao O$_{\text{SMOW}}$ of seawater has been fairly constant throughout the Phanerozoic, at around $-1\permil$. Several studies have suggested that high-temperature alteration of the seafloor acts as a buffer for the \deltao O of seawater \citep{Muehlenbachs1976,Gregory1981,Gregory1991,Muehlenbachs1998}, but a more recent model implied that this buffer may not be strong enough to overcome slow changes in the \deltao O of seawater \citep{Jaffres2007}. If this buffer cannot explain the relative constancy of seawater \deltao O values through time, then the two competing controls on seawater \deltao O, alteration of rocks at high and low temperatures, must have remained in balance over the last 500 million years. This suggests that the magnitudes of high-temperature hydrothermal alteration and low-temperature weathering have not changed substantially, or that any changes in the magnitude of one process are tied through an as-yet unidentified feedback loop to balancing changes in the other.

@@

\newpage

\section*{References}
\bibliographystyle{my-harv}
\bibliography{draft}

\begin{figure}[p]
\centering
\includegraphics[width=0.5\textwidth]{Fig_map.pdf}
\caption{Geologic map of the northern half of the island of Gotland, Sweden, modified from \cite{Calner2004a}. Approximate sample locations are marked by colored circles. Numbering corresponds to the sample location in Table \ref{results}.}
\label{map}
\end{figure}

\begin{figure}[htb]
\centering
\includegraphics[width=0.9\textwidth]{Fig_column-01.jpg}
\caption{On the left, a stratigraphic column showing the generalized lithstratigraphic units that were sampled for this study. On the right are the carbon and oxygen isotopic compositions of apparently well-preserved brachiopods, collected from the same section in Gotland. Isotopic compositions of brachiopods are modified from \cite{Bickert1997}.}
\label{column}
\end{figure}

\begin{sidewaysfigure}[htb]
\centering
\includegraphics[width=\textwidth]{Fig_section-01.jpg}
\caption{The \deltac C, \deltao O and $\Delta_{47}$ temperature of all the samples analyzed in this study, plotted against stratigraphic height. A height of zero meters is defined by the contact between the Lower Visby Beds and the Upper Visby Beds. Also included are data from Munnecke et al. (2003).}
\label{section}
\end{sidewaysfigure}

\begin{figure}[htb]
\centering
\includegraphics[width=\textwidth]{Fig_Brand-01.jpg}
\caption{Bulk manganese and strontium concentrations of the brachiopods, rugose corals, micrites and cements in this study. The brachiopods in this study have less Mn and more Sr than most of the brachiopods analyzed by Brand et al. (2012), which are also shown in this figure. Lines represent estimates by different authors of the range of pristine concentrations defined for brachiopods. Rugose corals generally have more Fe (not shown), more Mn and less Sr than brachiopods, giving them more altered trace metal signatures, while cements and micrites tend to have the most highly altered trace metal signatures.}
\label{Brand}
\end{figure}

\begin{figure}[htb]
\centering
\includegraphics[width=\textwidth]{Fig_metals-01.jpg}
\caption{Scatter plots showing the relationship between temperature and bulk trace metal concentration. The only statistically robust correlation ($\text{p}<0.05$ compared to zero correlation) is that between temperature and Sr concentration of rugose corals.}
\label{metals}
\end{figure}

\begin{figure}[htb]
\centering
\includegraphics[width=0.8\textwidth]{Fig_EBSD_coral-01.jpg}
\caption{Four different images of the same area of a rugose coral thin section, all at the same magnification. A) Reflected light. The line of dark pits are damaged caused by the E-probe electron beam. The dark triangle labelled ``RC'' was identified as a recrystallized domain of the coal septum. B) EBSD analysis. The recrystallized domain is clearly marked by many calcite crystals with the same orientation. The black scale bar represents 500 $\mu$m. C) Iron and D) manganese concentrations measured using the E-probe. The maps are qualitative, with warmer colors representing higher concentrations. The overlaid spots were measured quantitatively, and are labeled in units of ppm. For both metals, the recrystallized domain is indistinguishable from the rest of the septum.}
\label{EBSD_coral}
\end{figure}

\begin{figure}[htb]
\centering
\includegraphics[width=\textwidth]{Fig_EBSD_brach-01.jpg}
\caption{A) Transmitted polarized light image of a thin section of a brachiopod with $\Delta_{47}$ temperature 34\degrees C. The color overlay is an electron backscatter diffraction image of the same sample. The orientation of calcite crystals is indicated by color. B) Euler pole figure generated from the EBSD analysis.}
\label{EBSD_brach}
\end{figure}

\begin{figure}[htb]
\centering
\includegraphics[width=\textwidth]{Fig_flakes-01.jpg}
\caption{Scanning electron microscope images of flakes of brachiopod shell. Images were taken in secondary electron mode. Panels A--C show flakes that were judged to be pristine and are representative of the material used for analysis. Panel D shows a flake that was judged to be partially recrystallized and would not have been chosen for analysis. Clumped isotope temperatures of the samples for which representative flakes are shown are: A) 34 $\pm$ 2\degrees C; B) 35 $\pm$ 3\degrees C; C) 35 $\pm$ 3\degrees C.}
\label{flakes}
\end{figure}

\begin{figure}[htb]
\centering
\includegraphics[width=0.7\textwidth]{Fig_ellipses-01.jpg}
\caption{The $\Delta_{47}$ temperature and calculated water \deltao O for all samples analyzed in this study. Error ellipses represent one standard deviation of the mean of all measurement acquisitions for each sample.}
\label{ellipses}
\end{figure}

\begin{figure}[htb]
\centering
\includegraphics[width=0.5\textwidth]{Fig_comparisons-01.jpg}
\caption{Comparisons between the \deltac C, \deltao O and $\Delta_{47}$ temperature of brachiopod samples and adjacent micrite.}
\label{comparisons}
\end{figure}

\begin{figure}[htb]
\centering
\includegraphics[width=0.6\textwidth]{Fig_veizer-01.jpg}
\caption{In gray, the \deltao O$_{\text{PDB}}$ of apparently well-preserved skeletal calcites, compiled by \cite{Veizer1999}. In black are two different hypotheses for the trend in the \deltao O$_{\text{SMOW}}$ of the oceans through time.}
\label{Veizer}
\end{figure}

>>>>

\documentclass[5p, authoryear]{elsarticle}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{rotating}
\usepackage{multirow}
\usepackage{enumerate}
\usepackage{wasysym}
\usepackage{natbib}
\usepackage{lineno}
\usepackage{dblfloatfix}

\renewcommand{\topfraction}{0.99}
\renewcommand\bottomfraction{.99}
\renewcommand{\textfraction}{0.01}
\renewcommand{\floatpagefraction}{0.99}
\setcounter{totalnumber}{50}
\setcounter{topnumber}{50}
\setcounter{bottomnumber}{50}

@@

@@

\begin{figure*}[t]
\centering
\includegraphics[width=140mm]{Fig_01-01.png}
\caption{Geologic map of the northern half of the island of Gotland, Sweden, modified from \cite{Calner2004a}. Approximate sample locations are marked and labeled with location number, which corresponds to the sample locations in the supplementary table.}
\label{map}
\end{figure*}

@@

\begin{figure*}[t]
\centering
\includegraphics[width=140mm]{Fig_02-01.png}
\caption{Lithology and isotope stratigraphy of the sampled sequence. On the left, the stratigraphic column shows the generalized lithostratigraphic units that were sampled for this study. On the right are the carbon and oxygen isotopic compositions of apparently well-preserved brachiopods, collected from the same section in Gotland by \cite{Bickert1997}. Figure modified from \cite{Bickert1997}.}
\label{column}
\end{figure*}

@@

\begin{figure*}[b]
\centering
\includegraphics[width=\textwidth]{Fig_03_color-01.png}
\caption{The \deltac C, \deltao O and clumped isotope temperature of all the samples analyzed in this study, plotted against stratigraphic height. For the Lower \& Upper Visby and H\"{o}gklint Formations, a height of zero meters is defined by the contact between the Lower Visby Beds and the Upper Visby Beds. For the Slite Formation, a height of zero meters is defined as an arbitrary bed near the base of the outcrop. Brachiopod $\delta^{13}$C and $\delta^{18}$O measurements from \cite{Munnecke2003} are also included. Error bars represent the standard error of the mean.}
\label{section}
\end{figure*}

@@

\begin{figure}[h]
\centering
\includegraphics[width=90mm]{Fig_04_color-01.png}
\caption{Bulk manganese and strontium concentrations of the brachiopods, rugose corals, micrites and cements in this study. The brachiopods in this study have less Mn and more Sr than most of the brachiopods analyzed by \cite{Brand2012} which are also shown in this figure. Lines represent estimates by different authors of the range of pristine concentrations defined for brachiopods. Rugose corals generally have more Fe (not shown), more Mn and less Sr than brachiopods, giving them more altered trace metal signatures, while cements and micrites tend to have the most highly altered trace metal signatures. Figure modified from \cite{Brand2012}.}
\label{Brand}
\end{figure}

@@

\begin{figure*}[tb]
\centering
\includegraphics[width=\textwidth]{Fig_05_color-01.png}
\caption{Scatter plots showing the relationship between temperature and bulk trace metal concentrations. The negative correlation between temperature and the log of Sr concentration in rugose corals is statistically robust, and holds up when combined with the rest of the samples, but no other correlations passed our hypothesis tests. These plots demonstrate clear differences in trace metal content between the two fossil taxa and the later carbonate phases. Brachiopods tend to have the most pristine trace metal signatures, with less Mn and more Sr than the other samples. Micrites and cements tend to have the most altered trace metal signatures, with high Fe and Mn and low Sr concentrations. }
\label{metals}
\end{figure*}

@@

\begin{figure*}[tb]
\centering
\includegraphics[width=140mm]{Fig_06_color-01.png}
\caption{Calcite fabric and trace metal concentrations in a rugose coral. The four images show the same area of a rugose coral thin section, all at the same magnification. The black bar at the bottom of panel B is 500 $\mu$m long for scale. A) Transmitted light. The dark triangle labelled ``RC'' was identified as a recrystallized domain of the coral septum. Cement in the void space between coral septa appears as blocky calcite crystals. B) Electron backscatter diffraction (EBSD) analysis. The orientation of calcite crystals is indicated by different shades. The recrystallized domain is clearly marked by many calcite crystals with the same orientation. C) Iron and D) manganese concentrations measured using the E-probe. The maps are qualitative, with darker shades representing higher concentrations. For both metals, the recrystallized domain is indistinguishable from the rest of the septum. The black lines show the location of an array of 62 spots where trace metal concentrations were measured quantitatively. The concentration of Fe and Mn at these spots is plotted to the right of the respective map.}
\label{EBSD_coral}
\end{figure*}

@@

\begin{figure*}[tb]
\centering
\includegraphics[width=\textwidth]{Fig_07_color-01.png}
\caption{Calcite fabric in a brachiopod. A) Transmitted polarized light image of a thin section of brachiopod G3 LVF B1. This sample has a clumped isotope temperature of 35\degrees C. The boxed overlay is an electron backscatter diffraction (EBSD) image of the same sample. The orientation of calcite crystals is indicated by different shades. B) Euler pole figure generated from the EBSD analysis. The central band represents the preferential orientation of calcite crystals parallel to the outer surface of the brachiopod shell.}
\label{EBSD_brach}
\end{figure*}

@@

\begin{figure}[h]
\centering
\includegraphics[width=90mm]{Fig_08-01.png}
\caption{Scanning electron microscope images of flakes of brachiopod shell. Images were taken in secondary electron mode. Panels A--C show flakes that were judged to be pristine and are representative of the material used for analysis. Panel D shows a flake that was judged to be partially recrystallized and would not have been chosen for analysis. Clumped isotope temperatures of the samples for which representative flakes are shown are: A) 35 $\pm$ 2\degrees C; B) 33 $\pm$ 2\degrees C; C) 35 $\pm$ 1\degrees C.}
\label{flakes}
\end{figure}

@@

\begin{figure}[h]
\centering
\includegraphics[width=90mm]{Fig_09-01.png}
\caption{The clumped isotope temperature and calculated water \deltao O$_{\text{VSMOW}}$ for all samples analyzed in this study. Error ellipses represent one standard deviation of the mean of all measurement acquisitions for each sample.}
\label{ellipses}
\end{figure}

@@

\begin{figure}[h]
\centering
\includegraphics[width=90mm]{Fig_10-01.png}
\caption{Comparisons between the \deltac C, \deltao O and clumped isotope temperature of brachiopod samples and adjacent micrite.}
\label{comparisons}
\end{figure}

@@

Our results, along with those of previous clumped isotope studies \citep{Came2007,Dennis2013,Finnegan2011}, provide consistent explanations for the low \deltao O values of early Paleozoic carbonates (Figure \ref{Veizer}). Carbonate clumped isotopes constrain the contribution of a change in seawater \deltao O to only a few permil of the observed trend, ruling out a value of $-8\permil$ for the \deltao O$_{\text{SMOW}}$ of early Paleozoic oceans. Our results also serve as a reminder that many of the fossils preserved in the geologic record lived in tropical oceans. The \deltao O of carbonate in equilibrium with tropical ocean water should not be compared to the \deltao O of carbonate in equilibrium with water of average modern sea surface temperature, because tropical oceans are much warmer than the average sea surface temperature. Clumped isotopes additionally demonstrate that extremely hot ($>50$\degrees C) ocean temperatures cannot be the cause of the depleted \deltao O values of early Paleozoic carbonates. Finally, clumped isotopes, as a highly sensitive indicator of diagenetic alteration, demonstrate that even among samples with original calcite fabric and pristine trace metal signatures, there is a large range in preservation quality. This means that the average fossil calcite in almost any sample suite is likely altered from its primary composition, and that diagenetic alteration is a key contributor to the low \deltao O values of early Paleozoic carbonates.

@@

The results from this study join those of previous clumped isotope studies in implying that the \deltao O$_{\text{SMOW}}$ of seawater has been fairly constant throughout the Phanerozoic, at around $-1\permil$ (Figure \ref{Veizer}). Several studies have suggested that high-temperature alteration of the seafloor acts as a buffer for the \deltao O of seawater \citep{Muehlenbachs1976,Gregory1981,Gregory1991,Muehlenbachs1998}, but a more recent model implied that this buffer may not be strong enough to overcome slow changes in the \deltao O of seawater \citep{Jaffres2007}. If this buffer cannot explain the relative constancy of seawater \deltao O values through time, then the two competing controls on seawater \deltao O, alteration of rocks at high and low temperatures, must have remained in balance over the last 500 million years. This suggests that the magnitudes of high-temperature hydrothermal alteration and low-temperature weathering have not changed substantially, or that any changes in the magnitude of one process are tied through an as-yet unidentified feedback loop to balancing changes in the other.

\begin{figure}[h]
\centering
\includegraphics[width=90mm]{Fig_11-01.png}
\caption{Implications for the constancy of seawater $\delta^{18}$O through time. In gray, the \deltao O$_{\text{VPDB}}$ of apparently well-preserved skeletal calcites, compiled by \cite{Veizer1999}. The black solid line represents the estimated \deltao O$_{\text{VSMOW}}$ of the oceans through time based on the calcite values. The two black dashed lines represent alternative hypotheses for the value of seawater \deltao O$_{\text{VSMOW}}$ through time. All three lines ignore short-term perturbations  to seawater $\delta^{18}$O caused by glaciations. Finally, the black shapes are estimates of seawater \deltao O from clumped isotopes, measured at four different times by three separate studies. These clumped isotope estimates suggest that seawater \deltao O has not changed by more than 2$\permil$ in the last 430 million years.}
\label{Veizer}
\end{figure}

@@

\nocite{*}

\section*{References}
\bibliographystyle{my-geochim}
\bibliography{draft}
